---
title: "Immune_Bulk_RNAseq_Analysis"
output: rmarkdown::github_document
---

## Objectives

### Gene expression profiling by RNAseq in melanoma cell lines using data from the journal article "Conserved Interferon-g Signaling Drives Clinical Response to Immune Checkpoint Blockade Therapy in Melanoma" 

### Grasso et al., 2020, Cancer Cell 38, 500–515

### In this study the authors have discovered from an immunotherapy cohort that response to immune checkpoint blockade therapy is associated with T-cell infiltration and interferon gamma (IFNG). They modelled the effect of IFNG on human melanoma cell lines. 

### They have treated 58 cell lines with IFNG for 6 hours. Among these, some cell lines included mutations in JAK1/JAK2/B2M which might affect the downstream of IFNG pathway consequently the response to IFNG. They conducted RNA-seq before (control) and after IFNG treatment samples to find the transcriptomic changes. 

## Data

### Raw gene count data is accessible from GEO with the following accession code:GSE154996

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
pacman::p_load(tidyverse, readxl, openxlsx, janitor, ggrepel, ggplot2,
               edgeR, DESeq2, EnhancedVolcano, pheatmap,sjmisc,stringr,ComplexHeatmap,
               clusterProfiler,org.Hs.eg.db, AnnotationDbi, conflicted,readxl,gprofiler2)
conflict_prefer("filter", "dplyr") ## IMPORTANT
conflict_prefer("select", "dplyr")
conflict_prefer("intersect", "dplyr")
conflict_prefer("setdiff", "dplyr")

rm(list=ls())
```

## Read in count and filtering of genes with no expression

```{r, results='hide'}
counts <- read_delim("./data/GSE154996_Raw_Gene_Counts_Matrix.txt") %>% 
  clean_names() 

counts <- counts %>%
  clean_names() %>%
  select(gene_id, ends_with("_counts")) %>% 
  column_to_rownames("gene_id") 


counts <- counts %>% 
  rename_with(~gsub("_counts", "", .))

n_expressed <- rowSums(counts > 0)
sum(n_expressed==0) #997 genes no expression
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
ggplot(data.frame(n_expressed), aes(x=n_expressed)) +
  geom_bar() +
  labs(x="Number of samples expressed in", y="Number of genes") +
  theme_bw(base_size = 20)

```

```{r}
dim(counts) # 19871   116
counts <- counts[ n_expressed > 0, ] 
dim(counts) #18874   116 (997 genes that are not express were removed)

```
## GENE ANNOTATION

```{r}
gene_ann <- data.frame(SYMBOL=rownames(counts))  
head(gene_ann, 3)
map <- clusterProfiler::bitr(gene_ann$SYMBOL,
            fromType = "SYMBOL",
            toType = c("ENTREZID", "GENENAME", "GENETYPE"), 
            OrgDb = "org.Hs.eg.db")

tb <- table(map$SYMBOL) 
dup <- names(which(tb > 1)) 
map %>% filter( is.element(SYMBOL, dup) )
map <- map %>%
  filter( !is.element(SYMBOL, dup) ) 
dim(map)
gene_ann <- left_join(gene_ann, map)              

rownames(gene_ann) <- rownames(counts) 

head(gene_ann)

rm(tb, dup, map)

dim(gene_ann) # 18874    4

```

## Sample annotation

```{r}
sample_ann <- colnames(counts) %>% 
  as.data.frame() %>% 
  setNames("Exp_info") %>% 
  mutate(temp =  sub("_ko_[^_]+$", "", Exp_info),
         Cell_line = sub("_[^_]+$", "", temp), 
         Mutation = ifelse(grepl("jak1_ko", Exp_info),"jak1_ko", 
                           ifelse(grepl("jak2_ko", Exp_info),"jak2_ko",
                                  ifelse(grepl("b2m_ko", Exp_info),"b2m_ko","wt"))),
         Treatment =  str_sub(Exp_info,-4,-1), 
         Treatment =  factor(Treatment, levels=c("ctrl", "expt"))) %>% 
  select(-temp) %>% 
  column_to_rownames('Exp_info') 

sample_ann %>% tabyl(Treatment)

write.csv(sample_ann,"./data/sample_ann.csv", row.names = TRUE)

```
## Q1. Generate DESeq2 object using column/count data and	make a PCA plot using normalised count data.
## Are control and treatment samples separated on the PCA plot?

```{r}
# Check the order of samples in count and sample annotation

identical( colnames(counts), rownames(sample_ann) ) # identical order

identical( rownames(counts), rownames(gene_ann) ) # identical order

dds <- DESeqDataSetFromMatrix(countData = round(counts),
                              colData = sample_ann,
                              rowData = gene_ann,
                              design = ~ -1 + Treatment + Mutation + Cell_line) 
assay(dds)[ 1:3, 1:5]

cpm(dds)[1:3, 1:5]

cpm(dds) %>% colSums() %>% head()

colData(dds)[1:3, ]

rowData(dds)[1:3, ]

dim(dds) #18874   116

save(dds, file="./data/data_clean.rda", compress=TRUE)


```
## VST expression (Variable stabilizing)

```{r}

vst_expr <- vst(dds, blind=FALSE)
norm_count <- assay(vst_expr)

write.csv(norm_count,"./results/normalized_counts.csv", row.names = TRUE) 

```

## PCA Plot (By Treatment)

```{r}

plotPCA(vst_expr, intgroup="Treatment") + 
  theme_bw(base_size=20)

```
## PCA Plot (By Mutation)

```{r}

plotPCA(vst_expr, intgroup="Mutation") + # Specify color by Treatment as specified in Design
  theme_bw(base_size=20)

```
## Q2. Find differentially expressed genes (DEGs) between pre and post treatment  ## 
## samples and compare list of up and down-regulated genes with author's list   ##

```{r}

# Differential Gene Expression

gene_ann <- rowData(dds) %>%
  data.frame() %>%
  select(SYMBOL, ENTREZID)

head(gene_ann, 3)

dds <- DESeq(dds)

resultsNames(dds)

# Extract results for ctrl vs. expt 
res <- results(dds, contrast=c("Treatment", "expt", "ctrl")) %>% data.frame() 

res <- res %>%
  rownames_to_column("gene") %>%
  select(SYMBOL=gene, baseMean, LFC=log2FoldChange, p=pvalue, FDR=padj) %>%
  filter(!is.na(LFC), !is.na(FDR)) %>%
  arrange(p) %>%
  left_join(gene_ann) %>% 
  mutate( sig = sign(LFC) * (abs(LFC) > log2(1.5) & FDR < 0.05) )

res %>% tabyl(sig) # (downregulated: 41, upregulated: 537)

openxlsx::write.xlsx( list(res=res),file="./results/expt_up_deg_L2FC_q_filter.xlsx") 

```
## compare list of up and down-regulated genes with author's list


```{r, warning=FALSE}

article_selected <- read_excel("./data/mmc3.xlsx", skip = 1) %>%
  select("GeneId") %>% 
  pull()

res_selected <- res %>% 
  filter(sig != 0) %>%
  select("SYMBOL") %>% 
  pull()

library(VennDiagram)
# Venn Diagram to show intersections of two vectors
#svg(file="./results/venn.svg")
v2 <- venn.diagram( x = list(article_selected, res_selected),
                    category.names = c("article_selected", "res_selected"),
                    filename = NULL)
grid.newpage()
grid.draw(v2)
#dev.off()

intersected_gene <- Reduce(intersect, list(article_selected,res_selected))
openxlsx::write.xlsx( list(intersected_gene=intersected_gene),file="./results/intersected_gene.xlsx") 

res_gene_only <- Reduce(setdiff, list(res_selected, article_selected))
openxlsx::write.xlsx( list(res_gene_only=res_gene_only),file="./results/res_gene_only.xlsx") 

article_gene_only <- Reduce(setdiff, list(article_selected, res_selected)) %>% 
  as.data.frame() %>% 
  pull()
```

## Getting the list of excluded genes in results vs. authors' list
### Specify the filtering condition of DEGs by (abs(LFC) > log2(1.5) & FDR < 0.05)


```{r, warning=FALSE}

article_selected <- read_excel("./data/mmc3.xlsx", skip = 1)

library(data.table)  

article_selected_fc <- article_selected[article_selected$GeneId %in% article_gene_only,] %>% 
                      clean_names() %>% 
                      t() %>% 
                      as.data.frame() 
                      
article_selected_fc <- setDT(article_selected_fc, keep.rownames = "cell_line") %>%  
                       row_to_names(row_number = 1) %>%
                       mutate(cell_line =  sub("_log2fold_[^_]+$", "", gene_id)) %>% 
                       select(c(cell_line, PHF20, SLC26A2, KANK2))   
  
openxlsx::write.xlsx( list(article_selected_fc=article_selected_fc),file="./results/article_selected_fc.xlsx")

```

## Getting the list of intersected genes in our results vs. authors' list

```{r, warning=FALSE}

article_selected <- read_excel("./data/mmc3.xlsx", skip = 1)

intersected_selected_fc <- article_selected[article_selected$GeneId %in% intersected_gene,] %>% 
                           clean_names() %>% 
                           t() %>% 
                           as.data.frame() 

intersected_selected_fc <- setDT(intersected_selected_fc, keep.rownames = "cell_line") %>%  
  row_to_names(row_number = 1) %>%
  mutate(cell_line =  sub("_log2fold_[^_]+$", "", gene_id)) %>% 
  select(c(cell_line, IRF1, STAT3, TRAFD1))   

openxlsx::write.xlsx( list(intersected_selected_fc=intersected_selected_fc),file="./results/intersected_selected_fc.xlsx")

```


## Q3. Make a volcano plot of up and down-regulated genes.

```{r}

# Volcano Plot
# Visualize the biological (log fold change) vs. statistical significance (FDR)

#svg(file="./results/upregulated_genes_treatment.svg")
EnhancedVolcano(res, # results file (non-normalized)
                lab = res$SYMBOL, x = "LFC", y = "FDR",
                FCcutoff = log2(2), pCutoff = 0.05,
                title = "Upregulated genes with IFN-g treatment", subtitle = NULL,
                ylab = bquote(~-Log[10] ~ italic(FDR)), legendPosition = "none")
#dev.off()

```

## Q4. What are the top up and down regulated pathways?
## Reproduce similar Enrichment Plots in Fig. S9. 

```{r}

# GSEA with Hallmark

res <- read_excel("./results/expt_up_deg_L2FC_q_filter.xlsx")
ranked <- res %>% 
  arrange( desc(LFC) ) %>% # can also consider p-value for ranking
  select(ENTREZID, LFC) %>%
  na.omit() %>%
  deframe()

length(ranked)

# Next, prepare the Hallmark database

hallmark <- msigdbr::msigdbr(species = "human", category = "H") %>%
  select(gs_name, entrez_gene) %>%
  unique() %>%
  mutate( gs_name = gsub("HALLMARK_", "", gs_name) )

# Now run GSEA on Hallmark and look at the top few significant pathways:

gsea_hallmark <- GSEA(gene = ranked, TERM2GENE = hallmark)

gsea_hallmark <- setReadable(gsea_hallmark, OrgDb=org.Hs.eg.db, keyType="ENTREZID")

gsea_hallmark@result %>%
  filter(p.adjust < 1e-4) %>%
  arrange(NES) %>% 
  select(setSize, NES, p.adjust)

#svg(file="./results/Dot_blot_HM.svg")
dotplot(gsea_hallmark, showCategory=20, split=".sign") + labs(title="GSEA_Hallmark") + facet_grid(.~.sign)
#dev.off()

```

## GSEA Plot (INTERFERON_GAMMA_RESPONSE)

```{r, results='hide'}
keep <- gsea_hallmark@result %>% 
  filter(ID=="INTERFERON_GAMMA_RESPONSE") %>% 
  pull(core_enrichment) %>% 
  str_split(pattern="\\/") %>% 
  unlist() %>% 
  rev()          

keep <- data.frame(SYMBOL=keep)

left_join(keep, res)

```
```{r}
# Visualize the enrichment for this pathway

#svg(file="./results/Hallmark INTERFERON_GAMMA_RESPONSE.svg")
gseaplot(gsea_hallmark, geneSetID = "INTERFERON_GAMMA_RESPONSE", title = "Enrichment plot: \n HALLMARK_INTERFERON_GAMMA_RESPONSE")
#dev.off()
```



## GSEA-GO

```{r, results='hide'}

#Running GSEA-GO##
res_up_GO_BP <-gseGO(ranked, 'org.Hs.eg.db', ont="BP", pvalueCutoff=0.01) 

res_up_GO_BP <- setReadable(res_up_GO_BP, OrgDb=org.Hs.eg.db, keyType="ENTREZID")
## Pull out the top ranking terms. Add geneID to see the genes involved.
res_up_GO_BP@result %>%
  filter(p.adjust < 1e-5) %>%
  select(Description, setSize, NES, p.adjust)

```

```{r}
#svg(file="./results/Dot_blot_HM.svg")
dotplot(res_up_GO_BP, showCategory=5, split=".sign") + labs(title="GSEA_GO_BP") + facet_grid(.~.sign)
#dev.off()
```


## GSEA-GO (GO_RESPONSE_TO_INTERFERON_GAMMA)

```{r, results='hide'}

keep2 <- res_up_GO_BP@result %>% 
  filter(ID=="GO:0034341") %>% 
  pull(core_enrichment) %>% 
  str_split(pattern="\\/") %>% 
  unlist() %>% 
  rev()          

keep2 <- data.frame(SYMBOL=keep2)

left_join(keep2, res)

```

```{r}
#svg(file="./results/GO_Response to interferon-gamma.svg")
gseaplot(res_up_GO_BP, geneSetID = "GO:0034341", title = "Enrichment plot: \n GO_RESPONSE_TO_INTERFERON_GAMMA")
#dev.off()
```


## The top upregulated pathways are: INTERFERON_GAMMA_RESPONSE, INTERFERON_ALPHA_RESPONSE, IL6_JAK_STAT3_SIGNALING, ALLOGRAFT_REJECTION, INFLAMMATORY_RESPONSE, IL2_STAT5_SIGNALING, COMPLEMENT, TNFA_SIGNALING_VIA_NFKB

```{r}

#svg(file="./results/Dot_blot.svg")
g1 <- dotplot(gsea_hallmark, showCategory=20, split=".sign") + labs(title="GSEA_Hallmark") + facet_grid(.~.sign)
g2 <- dotplot(res_up_GO_BP, showCategory=5, split=".sign") + labs(title="GSEA_GO_BP") + facet_grid(.~.sign)
cowplot::plot_grid(g1, g2, nrow=1)
#dev.off()

```

## Q5. Reproduce similar heatmap in Fig. 3A. (Heatmap of log2 fold change)

## Heatmap
### Read in the csv file containing the genes to plot in heatmap
### Import normalise count

```{r, results='hide'}

norm_count2 <- read.csv("./results/normalized_counts.csv")

colnames(norm_count2)[1] ="gene_id"

head(norm_count2)

```


## Import gene list

```{r}

gene_list <- read_csv("./data/Heatmap_genes.csv") %>% pull("gene_id")

filterlist <- norm_count2 %>% 
              filter(gene_id %in% gene_list) %>% 
              column_to_rownames("gene_id") %>% 
              t() %>% 
              as.data.frame() %>% 
              rownames_to_column("sample")  
              
openxlsx::write.xlsx( list(filterlist=filterlist),file="./results/filterlist_vst.xlsx")

```

## Generating log2FC

```{r}


DF1 <- filterlist %>% 
       filter(grepl("_expt", sample)) %>% 
       mutate(sample = gsub("_expt", "", sample)) %>% 
       column_to_rownames("sample")

DF2 <- filterlist %>% 
       filter(grepl("_ctrl", sample)) %>% 
       mutate(sample = gsub("_ctrl", "", sample)) %>% 
       column_to_rownames("sample")

```


## Check order of column name

```{r}

identical( rownames(DF1), rownames(DF2) )  

DF1 = DF1[rownames(DF2),] 

identical( rownames(DF1), rownames(DF2) )

```


## subtraction

```{r}

L2FC <- DF1 -DF2
min(L2FC)
max(L2FC)
L2FC_mat <- as.matrix(L2FC) %>%  t()

L2FC_mat = L2FC_mat[gene_list,]

pacman::p_load(circlize)

#svg(file="./results/Heatmap.svg", width=10, height=9)
Heatmap(L2FC_mat,cluster_rows = F,
        column_names_gp = grid::gpar(fontsize = 8),column_title = "Samples", row_title = "Gene_list",
        row_names_gp = grid::gpar(fontsize = 8), name = "Scale",
        col = colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))) 
#dev.off()

```
